Twenty years of dynamic occupancy models: a review of applications and a look towards the future

### 2024-09-20 _manuscript/Manuscript.qmd

The largest differences between the frequentist and Bayesian models in our sample lie in their approaches to model selection. Where 94% of frequentist models perform some manner of model selection to determine covariate inclusions for their models, only 36% of Bayesian models do so, with the majority instead fitting a single model defined *a priori*. Methods used for model selection vary considerably both within and between the two modelling frameworks. For frequentist models, the most popular and conventional approach to model selection as used in 43% of articles involves the creation of a fully pre-defined model set containing some number of hypothesis candidate models, where the best model is selected according to the lowest AIC score. The next most popular method in frequentist studies is to use procedural model selection methods (39% of articles), where the structure for each model parameter is fit one-by-one. For example, this protocol might first identify the best structure for detection probability before moving on to initial occupancy and so on. The remainder of frequentist studies (9%) use a variety of approaches, such as fitting simpler models like single season occupancy models to identify the most informative terms to use in a DOM. Across all frequentist implementations, 71% of articles summaruise a final subset of models using multi-model inference and model-averaging with AIC weights [@burnham2004].

Those Bayesian models which *do* perform model selection tend to take various approaches, with largely idiosyncratic methods across these studies. While direct comparison of model fit in Bayesian methods is rare, it is feasible — @urban2023 identifies the best model from a Bayesian candidate set using the predictive performance on both in and out-of-sample validation data. Another approach used by @cook2022 fit a global model including all covariates before removing each covariate where the 95% credible interval of the posterior distribution overlapped zero and refitting the model. @ahumada2013 takes a hybrid approach, in which model selection is conducted by a procedural method in the frequentist framework before refitting the best structure in a Bayesian model.

@@

The model selection methods identified in our review are designed only to identify the best model of those considered, and do not necessarily guarantee that selected models are well-fit to the data. Assessment of the performance of DOMs generally requires additional steps, and the best ways to achieve this are not necessarily clear. There is no broadly-accepted goodness-of-fit test for dynamic occupancy models, although @mackenzie2004 describes an approach for single-season occupancy models using a parametric bootstrap which can been extended to DOMs; this test is implemented in the *AICcModAvg* and *unmarked* R packages. @kéry2021 describe the test and present an alternative based on separately assessing fit to static and dynamic components of the model. In Bayesian implementations, posterior predictive checks offer means to assess model fit [@gelman2014]. @broms2016 discusses possibilities for model evaluation in the Bayesian context focusing on single-season multi-species occupancy models but extensions of their approach may also be applicable to the DOM. As in other hierarchical models, model evaluation for DOMs can in general can be difficult and somewhat uncertain compared to other model types, as the primary response variable of interest (species occupancy) is a latent variable where the true state is generally not known. Predictive performance evaluation is thus typically based on *observed* occupancy data, where a DOM is used to simulate detection histories to be compared with field survey results.

@@

Studies in our sample were classified according to their objectives as stated by authors to explore the use-cases which the DOM has been applied to (@fig-objectives). These objective categories were non-exclusive, and many studies fell into multiple categories; the frequency at which studies pursued each of these objectives is shown in @fig-objectives A. Across our study period DOMs have been frequently employed to estimate trends in species occupancy, to explore relationships between environmental factors and occupancy, and, increasingly often, to make predictions of occupancy spatially to unsurveyed locations or temporally to hypothetical future conditions. Within each of these broad categories lies even more variation in objectives, emphasising the DOM's flexibility as a tool for making ecological inference in diverse contexts.

@@

The objectives of studies fitting DOMs strongly inform approaches to the model building process, particularly with respect to covariate selection. @fig-objectives B shows the quantity of covariates considered for models applied to each of our objective categories, and the differences between objectives are apparent. Not unexpectedly, articles which focused on describing new methods for DOMs had fewer covariates than those studies focused on more applied objectives. However, differences persist between those articles observing trends, identifying trends in occupancy, testing relationships, and making predictions.

@@

Approaches to building any type of model will necessarily depend on the possibilities of the data at hand and on the priorities of the model-builder. This precludes any prescription of the 'best' way to build a model, however, there are still important discussions to be had on decisions made in the modelling process. One aspect of fitting DOMs meriting broader discussion revolves around 'model complexity,' and how much must be incorporated into models to reliable model species occupancy under different contexts and use cases. Complexity is a broad term which encompasses many aspects of a model [@merow2014], and opinions on simplicity versus complexity in ecological models can be divisive. Where some advocate for the simplest possible models, arguing that they are most generaliseable; others insist that overly-simple models cannot adequately represent the most important drivers in a system [@evans2013; @lonergan2014]. By their nature DOMs are somewhat more complex than simpler models for studying occupancy thanks to their hierarchical structure, attributes which are necessary to control for detectability and to capture occupancy dynamics. Within this structure, however, further complexity is to some degree up to the modeller: one can choose how many covariates to consider for inclusion on the various parameters, and how to represent the nature of the relationship between those covariates and parameters. Research from SDMs indicates that allowing for more complex relationships can improve model performance in predicting occupancy [@valavi2023], an increasingly popular use-case for DOMs. Within the DOM framework, there are exciting developments on that front — @joseph2020 presents a novel neural-network occupancy model which allows for exponentially higher levels of complexity and may offer improved performance for prediction-oriented studies.

Generally speaking, covariate selection seems to be a particularly important area for further investigations into building DOMs. In our review, we see little consensus around which approaches are most applicable for any given use case, and existing work on covariate selection in DOMs raises concerns on whether common methods always produce the most suitable models. This is true for both frequentist and Bayesian implementations, and comparative research on covariate selection under both frameworks may help to inform model users on which method may be most appropriate for their use-cases. In a similar vein, the low number of articles in our review which conducted model evaluation or assessed model fit raises different questions. While the appropriate method of model evaluation may depend on data availability and research objectives, assessing models by some method is generally important to understand how reliable model outputs may be [@araújo2005; @guisan2005]. Existing uncertainties around whether current methods are suitable for the task surely discourage users from calculating these metrics, and further research is needed to establish trusted practices for assessing the quality of DOMs.

These priorities are particularly important given the frequent applied objectives of DOM users, tackling challenges which include assessing critically endangered species [@carvalho2023], guiding public health management of disease vectors [@moreira2016], and tracking rapidly developing biological invasions [@wood2020]. DOMs are well suited to these situations, and it is to be expected that as these types of applications are more commonly attempted understanding the sensitivity of model outputs to decisions made in the model fitting process becomes increasingly important. In the two decades since the publication of @mackenzie2003 the dynamic occupancy model has become a widely used tool for ecological inference, with numerous extensions to the modelling framework further broadening the scope of questions and use-cases for which it may be applied. Given their increasing popularity, further research and guidelines around issues of model building may help to make the DOM format more accessible to newcomers and support confidence in interpretation of models. Parallels to existing work in the SDM literature on guidelines for reporting and modelling standards could be valuable contributions in achieving these aims [@zurell2020; @araújo2019] while specifically addressing model attributes unique to DOMs.

>>>>

The largest differences between the frequentist and Bayesian models in our sample lie in their approaches to model selection. Where 94% of frequentist models perform some manner of model selection to determine covariate inclusions for their models, only 36% of Bayesian models do so, with the majority instead fitting a single model defined *a priori*. Methods used for model selection vary considerably both within and between the two modelling frameworks (@tbl-modelling). For frequentist models, the most popular and conventional approach to model selection as used in 43% of articles involves the creation of a fully pre-defined model set containing some number of hypothesis candidate models, where the best model is selected according to the lowest AIC score. The next most popular method in frequentist studies is to use procedural model selection methods (39% of articles), where the structure for each model parameter is fit one-by-one. For example, this protocol might first identify the best structure for detection probability before moving on to initial occupancy and so on. The remainder of frequentist studies (9%) use a variety of approaches, such as fitting simpler models like single season occupancy models to identify the most informative terms to use in a DOM. Across all frequentist implementations, 71% of articles summarise a final subset of models using multi-model inference and model-averaging with AIC weights [@burnham2004].

Those Bayesian models which *do* perform model selection tend to take various approaches, with largely idiosyncratic methods across these studies. While direct comparison of model fit is rare in Bayesian methods, it is feasible — @urban2023 identifies the best model from a Bayesian candidate set using the predictive performance on both in and out-of-sample validation data. Another approach used by @cook2022 fit a global model including all covariates before removing each covariate where the 95% credible interval of the posterior distribution overlapped zero and refitting the model. @ahumada2013 takes a hybrid approach, in which model selection is conducted by a procedural method in the frequentist framework before refitting the best structure in a Bayesian model.

@@

The model selection methods identified in our review are designed only to identify the best model of those considered, and do not necessarily guarantee that selected models are well-fit to the data. Assessment of the performance of DOMs generally requires additional steps, and the best ways to achieve this are not yet clear. There is no broadly-accepted goodness-of-fit test for dynamic occupancy models, although @mackenzie2004 describes an approach for single-season occupancy models using a parametric bootstrap which can been extended to DOMs; this test is implemented in the *AICcModAvg* and *unmarked* R packages. @kéry2021 describe the test and present an alternative based on separately assessing fit to static and dynamic components of the model. In Bayesian implementations, posterior predictive checks offer means to assess model fit [@gelman2014]. @broms2016 discusses possibilities for model evaluation in the Bayesian context focusing on single-season multi-species occupancy models but extensions of their approach may also be applicable to the DOM. As in other hierarchical models, model evaluation for DOMs can in general can be difficult and somewhat uncertain compared to other model types, as the primary response variable of interest (species occupancy) is a latent variable where the true state is generally not known. Predictive performance evaluation is thus typically based on *observed* occupancy data, where a DOM is used to simulate detection histories to be compared with field survey results.

@@

Studies in our sample were classified according to their objectives as stated by authors to explore the use-cases which the DOM has been applied to (@fig-objectives). These objective categories were non-exclusive, and many studies fell into multiple categories. Across the study period the DOM has been frequently employed to estimate trends in species occupancy, to explore relationships between environmental factors and occupancy, and, increasingly often, to make predictions of occupancy spatially to unsurveyed locations or temporally to hypothetical future conditions. Within each of these broad categories lies even more variation in objectives, emphasising the DOM's flexibility as a tool for making ecological inference in diverse contexts.

@@

The objectives of studies fitting a DOM strongly informs approaches to model building, particularly with respect to covariate selection. @fig-objectives B shows the quantity of covariates considered for models applied to each of our objective categories, and the differences between objectives are apparent. Not unexpectedly, articles which focused on describing new methods for DOMs had fewer covariates than those studies focused on more applied objectives. However, differences persist between those articles observing trends, identifying trends in occupancy, testing relationships, and making predictions.

@@

Approaches to building any type of model will necessarily depend on the possibilities of the data at hand and on the priorities of the model-builder. This precludes any prescription of the 'best' way to build a model, however, there are still important discussions to be had on decisions made in the modelling process. One aspect of fitting DOMs meriting broader discussion revolves around 'model complexity,' and how much must be incorporated into models to reliably model species occupancy under different contexts and use cases. Complexity is a broad term which encompasses many aspects of a model [@merow2014], and opinions on simplicity versus complexity in ecological models are diverse. Where some advocate for the simplest possible models, arguing that they are most generalisable; others insist that overly-simple models cannot adequately represent the most important drivers in a system [@evans2013; @lonergan2014]. By their nature the DOM is somewhat more complex than simpler models for studying occupancy thanks to their hierarchical structure, which is necessary to control for detectability and to capture occupancy dynamics. Within this structure, however, further complexity is to some degree up to the modeller: one can choose how many covariates to consider for inclusion on the various parameters, and how to represent the nature of the relationship between those covariates and parameters. Research from SDMs indicates that allowing for more complex relationships can improve model performance in predicting occupancy [@valavi2023], an increasingly popular use-case for the DOM. Within the DOM framework, there are promising developments on that front — for example, @joseph2020 presents a novel neural-network occupancy model which allows for exponentially higher levels of complexity and may offer improved performance for prediction-oriented studies.

Covariate selection seems to be a particularly important area for further investigations into building DOMs. In our review, we see little consensus around which approaches are most applicable for any given use case, and existing work on covariate selection in DOMs raises concerns on whether common methods always produce the most suitable models. This is true for both frequentist and Bayesian implementations, and comparative research on covariate selection under both frameworks may help to inform model users on which method is most appropriate for their use-cases. In a similar vein, the low number of articles in our review that conducted model evaluation or assessed model fit raises different questions. While the appropriate method of model evaluation may depend on data availability and research objectives, assessing models by some method is generally important to understand how reliable model outputs may be [@araújo2005; @guisan2005]. Existing uncertainties around whether current methods are suitable for the task likely discourage users from calculating these metrics, and further research is needed to establish trusted practices for assessing the quality of DOMs.

These priorities are particularly important given the frequent applied objectives of DOM users, tackling challenges which include assessing critically endangered species [@carvalho2023], guiding public health management of disease vectors [@moreira2016], and tracking rapidly developing biological invasions [@wood2020]. The DOM is well suited to these situations, and it is to be expected that as these types of applications are more commonly attempted understanding the sensitivity of model outputs to decisions made in the model fitting process becomes increasingly important. In the two decades since the publication of @mackenzie2003 the dynamic occupancy model has become a widely used tool for ecological inference, with numerous extensions to the modelling framework further broadening the scope of questions and use-cases for which it may be applied. Given their increasing popularity, further research and guidelines around issues of model building may help to make the DOM more accessible to newcomers and support confidence in interpretation of models. Parallels to existing work in the SDM literature on guidelines for reporting and modelling standards could be valuable contributions in achieving these aims [@zurell2020; @araújo2019] could provide useful examples in developing methods and guidelines specifically focussed on the DOM.
